Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-09 manuscript/manuscript.tex
minor intro edit

Despite widespread interest in determining the evolutionary factors affecting neutral diversity on both X and Y chromosomes \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of demography or selection on patterns of sex-linked diversity in more recently evolved sex chromosomes, such as in dioecious plants. This information is important, however, as it highlights the relevant time scales over which effects such as linked selection are likely to be important. In humans, estimates of Y-linked diversity are considerably lower than predicted under models of neutral evolution, with simulations showing that current levels of Y-diversity are consistent with a combination of very low $N_{m}$ and a history of strong purifying selection \citep{Wilsonsayres2014}. However, given that human sex chromosomes evolved from autosomes \textasciitilde 200 million years ago (MYA), is not immediately obvious whether such strong effects of purifying selection on diversity should also apply to sex chromosomes that evolved \textit{do novo} from autosomes over much more recent evolutionary time (e.g., within the last 20 MYA in the case of most dioecious plants \citep{charlesworth2015plant}). The diversity loss due to purifying selection might on the one hand be expected to be lower in such systems due recombination having been suppressed on Y-chromosome for a shorter length of time. Contrary to this intuition, however, simulations have shown that two of the key types if HRI - background selection and selective sweeps - are both expected to have the strongest effects during the earliest stages of sex chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}. Indeed, this is consistent with studies of Y-degeneration in young plant sex chromosomes that have found signs of rapid genetic degeneration \citep{hough2014,papadopulos2015,charlesworth2016plant}. In these cases, if the early stages of Y-chromosome evolution involved widespread Y-linked gene-inactivation (silencing), as predicted by \citep{orr1998adaptive}, and discovered on the neo-Y chromosome in \textit{Drosophila mrianada} \citep{zhou2012chromosome}, then it is possible that linked selection plays a less important role than expected in the early stages, with gene loss instead being primarily driven by gene-silencing followed by neutral genetic drift \citep{bachtrog2013NRG}.

>>>>

Despite widespread interest in determining the evolutionary factors affecting neutral diversity on both X and Y chromosomes \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of demography or selection on patterns of sex-linked diversity in more recently evolved sex chromosomes, such as in dioecious plants. This information is important, however, as it highlights the relevant time scales over which effects such as linked selection are likely to be important. In humans, estimates of Y-linked diversity are considerably lower than predicted under models of neutral evolution, with simulations showing that current levels of Y-diversity are consistent with a combination of very low $N_{m}$ and a history of strong purifying selection \citep{Wilsonsayres2014}. However, given that human sex chromosomes evolved from autosomes \textasciitilde 200 million years ago (MYA), is not immediately obvious whether such strong effects of purifying selection on diversity should also apply to sex chromosomes that evolved \textit{do novo} from autosomes over much more recent evolutionary time (e.g., within the last 20 MYA in the case of most dioecious plants \citep{charlesworth2015plant}). Y-linked diversity loss due to purifying selection might on the one hand be expected to be lower in such systems due to recombination having been suppressed for a shorter length of time. Contrary to this intuition, however, simulations have shown that two key types of HRI - background selection and selective sweeps - both have strongest effects during the earliest stages of sex chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}. Indeed, this is consistent with studies of Y-degeneration in young plant sex chromosomes that have found signs of rapid genetic degeneration \citep{hough2014,papadopulos2015,charlesworth2016plant}. In these cases, if the early stages of Y-chromosome evolution involved widespread Y-inactivation (gene silencing), as predicted by \citep{orr1998adaptive}, and discovered on the neo-Y in \textit{Drosophila mrianada} \citep{zhou2012chromosome}, then linked selection might play a less important role than expected, with the early stages of Y-degeneration being primarily driven by gene-silencing followed by neutral genetic drift \citep{bachtrog2013NRG}.
